Bandwidth-Aware Reconfigurable Memory Design with Hybrid Memory Technologies

### 2011-04-09 trunk/paper.tex

One potential bottleneck for chip-multiprocessor~(CMP) performance scaling is
the widening gap between the bandwidth demand created by processor cores and the
limited access bandwidth provided by off-chip main memory [15][16][6][20]. Such
limitation greatly affects the parallelism of memory demanding applications,
i.e., applications with a large working set, by consuming additional cycles on
off-chip memory access. In addition, even moderate memory demand applications
will be affected as the number of cores scales up [17]. The continuing shrink of
transistor density not only leads to increasingly powerful processors with a
large number of cores, but also puts much bandwidth pressure to off-chip main
memory. However, the bandwidth to the off-chip main memory does not improve much
compared to the processor core scaling. Potentially, these issues make the
applications running on the computing systems be bandwidth-bound at main memory.

While a number techniques can be found in today's systems and research work to
alleviate such bandwidth bottleneck, few can improve the system throughput in a
power and cost efficient manner. High performance computing machines such as
NVIDIA's Tesla~\cite{nvidia:tesla} rely on very high main memory bandwidths
(provided by GDDR memories) to feed the requirement from the processor. However,
GDDR memories run at higher clock rates and are more power hungry than
conventional DRAM modules. It is undesirable for either general purpose or high
performance computing systems to improve their computing performance by
sacrificing power efficiency. With the emerging 3D chip technology, caches can
be stacked on top of processor cores to provide high memory
bandwidth[12][21][22]. However, ...

Caching has long been employed as the most effective approach to reduce memory
access latency. Proper on-chip memory hierarchy design, however, can also help
alleviate the increasing bandwidth pressure of off-chip memory. In [dac.1] the
authors discuss extensively the problem of limited pin bandwidth to
multiprocessor systems. They have focused on program performance in
multiprocessor systems and make detailed decomposition of program execution
times. They conclude that even more complex on-chip cache structures would prove
to be costeffective, with the limited pin bandwidth severely restricting
performance increases. In [dac.2] the authors have carefully studied the
requirements for on-chip cache structures along with all sorts of optimization
techniques that come along with scaling of processor core numbers. Their study
shows that cache size need to grow much faster than processor core numbers to
compensate for the limited off-chip bandwidth. Because of that, the near future
processors need to allocate a huge percentage of chip area for caches, which
means much less core counts than expected. The study also shows that effective
bandwidth optimization techniques can help reduce cache size requirement and
thus help scaling processor cores. However, the effectiveness of caching in the
mid/large scale multiprocessor systems could be highly suboptimal due to
significant contention across the parallel tasks. The more bandwidth requirement
by a workload can be satisfied by the last level cache~(LLC), the less bandwidth
will be required from the off-chip main memory. Our memory technology
pre-exploration shows that the bandwidth provided by a memory decreases with the
increase of its capacity. Based on these observations, adding more levels of
cache can potentially fill in the bandwidth gap.

@@

A large body of recent research focuses on exploring new memory technologies to
trade off between latency, bandwidth, and cost. Wu et al. explored various
memory technologies - SRAM, eDRAM [23], magnetic RAM (MRAM) [12] and PCM to best
construct L3 caches in terms of performance and power [24]. The eDRAM and MRAM
technologies are evaluated as LLCs by stacking each memory on top of the
processor die. Both studies mainly focused on reducing the latency gap between
L2 cache~(LLC) and external memory, but do not examine the design from bandwidth
perspective.

Application behavior prediction is a critical component of reconfigurable
architectures. Zhou \emph{et al.} monitored memory access patterns and estimated
memory behavior of workloads for energy efficient memory allocation
[predictor.26]. Duesterwald et al. [10] describe different statistical and table
based predictors for within- and across-metric predictions of performance
monitoring information. They show that the table-based predictor generally
outperforms the other predictors they tested. Sarikaya et al. [19] describe an
optimal prediction technique based on a predictive least squares minimization.
Sarikaya \emph{et al.} showed the benefit of statistical metric modeling for
tracking varying pattern history lengths and modeling long term
patterns~\cite{Sarikaya:2011:RWBP}. However, ...

@@

% \begin{figure}[t]
% \centering
% \includegraphics[width=2.5in]{figures/GST}
% \caption{The schematic view of a PCRAM cell with MOSFET selector transistor (BL=Bitline, WL=Wordline, SL=Sourceline}
% \label{fig:GST}
% \end{figure}

>>>>

One potential bottleneck for chip-multiprocessor~(CMP) performance scaling is
the widening gap between the bandwidth demand created by processor cores and the
limited access bandwidth provided by off-chip main
memory~\cite{McKee:2004:RMW,Burger:1996:MBLFM,Rogers:2009:SBW}. Such limitation greatly affects
the parallelism of memory demanding applications, i.e., applications with a
large working set, by consuming additional cycles on off-chip memory access. In
addition, even moderate memory demand applications will be affected as the
number of cores scales up~\cite{Huh:2001:EDSFC}. The continuing shrink of
transistor density not only leads to increasingly powerful processors with a
large number of cores, but also puts much bandwidth pressure to off-chip main
memory. However, the bandwidth to the off-chip main memory does not improve much
compared to the processor core scaling. Potentially, these issues make the
applications running on the computing systems be bandwidth-bound at main memory.

While a number techniques can be found in today's systems and research work to
alleviate such bandwidth bottleneck, few can improve the system throughput in a
power and cost efficient manner. High performance computing machines such as
NVIDIA's Tesla~\cite{Lindholm:2008:Tesla} rely on very high main memory
bandwidths (provided by GDDR memories) to feed the requirement from the
processor. However, GDDR memories run at higher clock rates and are more power
hungry than conventional DRAM modules. It is undesirable for either general
purpose or high performance computing systems to improve their computing
performance by sacrificing power efficiency. With the emerging 3D chip
technology, caches can be stacked on top of processor cores to provide high
memory bandwidth~\cite{Sun:2009:E3DSL2}. However, ...

Caching has long been employed as the most effective approach to reduce memory
access latency. Proper on-chip memory hierarchy design, however, can also help
alleviate the increasing bandwidth pressure of off-chip memory. After an
extensive study the problem of limited pin bandwidth to multiprocessor systems,
Burger \emph{et al.} conclude that even more complex on-chip cache structures
would prove to be costeffective with the limited pin bandwidth severely
restricting performance increases~\cite{Burger:1996:MBLFM}. The requirements of
on-chip cache structures along with all sorts of optimization techniques that
come along with scaling of processor core numbers are carefully
explored~\cite{Rogers:2009:SBW}. The study shows that cache size need to grow
much faster than processor core numbers to compensate the limited off-chip
bandwidth. The study also shows that effective bandwidth optimization techniques
can help reduce cache size requirement and thus help scaling processor cores.
Consequently, it is critical to re-design the on-chip memory system, which
focuses on bandwidth optimization.

@@

A large body of recent research focuses on exploring new memory technologies to
trade off between latency, bandwidth, and cost. Wu et al. explored various
memory technologies - SRAM, eDRAM~\cite{Wu:2009:HCA}, and
STT-RAM~\cite{Sun:2009:NA3D} to best construct L3 caches in terms of performance
and power. The eDRAM and STT-RAM technologies are evaluated as LLCs by stacking
each memory on top of the processor die. Both studies mainly focused on reducing
the latency gap between L2 cache~(LLC) and external memory, but do not examine
the design from bandwidth perspective.

Application behavior prediction is a critical component of reconfigurable
architectures. Zhou \emph{et al.} monitored memory access patterns and estimated
memory behavior of workloads for energy efficient memory
allocation~\cite{Zhou:2004:DTP}. Duesterwald \emph{et al.} describe different
statistical and table based predictors for within- and across-metric predictions
of performance monitoring information~\cite{Duesterwald:2003:CPP}. They show
that the table-based predictor generally outperforms the other predictors they
tested. Sarikaya \emph{et al.}describe an optimal prediction technique based on
a predictive least squares minimization~\cite{Sarikaya:2007:PPB}. Sarikaya
\emph{et al.} showed the benefit of statistical metric modeling for tracking
varying pattern history lengths and modeling long term
patterns~\cite{Sarikaya:2010:RWB}. However, ...

@@

% \begin{figure}[t]
% \centering
% \includegraphics[width=2.5in]{figures/GST}
% \caption{The schematic view of a PCRAM cell with MOSFET selector transistor
% (BL=Bitline, WL=Wordline, SL=Sourceline}
% \label{fig:GST}
% \end{figure}
